Deterministic Bayesian inference methods for the Naomi model

### 2023-03-29 src/docs_11-04-23-explainer/11-04-23-explainer.Rmd
More on ELGM, clarify use of R-INLA copy, multiple link functions, moments and KS #24

* To explain what an extended latent Gaussian model (ELGM) is, we need a little more detail on LGMs
* In an LGM the conditional mean depends on exactly one structured additive predictor, specifically
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}

@@

* ELGM remove this requirement such that
$$
\mu_i = g(\eta_{\mathcal{J}_i})
$$
where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
* This allow for a higher degree of non-linearity in the model

@@

| | Things I thought made Naomi an ELGM | Does it actually? |
|-|---------------|--------------------|
|1| ANC offset from household survey | ? |
|2| Incidence depends on adult prevalence and coverage | ? |
|3| ART coverage and recent infection are products | ? |
|4| ART attendance uses a multinomial | ? |
|5| Aggregation of finer processes | ? |

* I will explain each of these, and whether or not they make Naomi an ELGM, in more detail in slides to follow
  * The notation may not have been introduced properly, but hopefully the gist will still make sense

@@

* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
\begin{align*}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
\end{align*}
* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian
* `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ but $A$ must be $n \times n$
* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM

@@

@@

<!-- # Moments -->

<!-- * Are the means and standard deviations the same? More generally the $t$th moment -->
<!-- $$ -->
<!-- \hat \mathbb{E}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t -->
<!-- $$ -->
<!-- * Won't do a great job if you're only looking at the first few moments $t = 1, 2$ -->

<!-- # Kolmogorov-Smirov test -->

<!-- * Compare $F_\bullet(\vartheta)$ to $F_\texttt{NUTS}(\vartheta)$ -->
<!-- $$ -->
<!-- D_\bullet = \sup_\vartheta | F_\texttt{NUTS}(\vartheta) - F_\bullet(\vartheta)|. -->
<!-- $$ -->
<!-- * A value $D_\bullet$ means there is at most a $100 \cdot D_\bullet$% difference between posterior densities -->

<!-- # -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."} -->
<!-- plot(1:10) -->
<!-- ``` -->

# Maximum mean discrepancy

# Pareto smoothed importance sampling

@@

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work, joint with Alex Stringer (Waterloo) and my PhD supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
* Let me know if you'd be up for being an early reader!
* Code for this project is at `athowes.github.io/elgm-inf`

>>>>

* Before defining extended latent Gaussian model (ELGM), first more detail on LGMs:
* Conditional mean depends on exactly one structured additive predictor
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$)

@@

* ELGM remove this requirement such that
$$
\mu_i = g_i(\eta_{\mathcal{J}_i})
$$
where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices
* Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$
  * $N_n < n$: more data points than structured additive predictors
  * $N_n = n$: as many data points as structured additive predictors (LGM case)
  * $N_n > n$: fewer data points than structured additive predictors
* The $g_i$ allow for a higher degree of non-linearity in the model

@@

| | Things I thought made Naomi an ELGM | Does it? |
|-|---------------|--------------------|
|1| ANC offset from household survey | ? |
|2| Incidence depends on adult prevalence and coverage | ? |
|3| ART coverage and recent infection are products | ? |
|4| ART attendance uses a multinomial | ? |
|5| Aggregation of finer processes | ? |
|6| Multiple link functions | ? |

* I will explain each of these^[The notation may not have been introduced properly, but hopefully the gist will still make sense], and whether or not they make Naomi an ELGM, in more detail in slides to follow

@@

* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
\begin{align*}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
\end{align*}
* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$
* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM

* Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right]
$$
\begin{pmatrix}
\eta^\star \\
\eta^\star
\end{pmatrix}
=
\begin{pmatrix}
1 & 1 \\
0 & 1
\end{pmatrix}
\begin{pmatrix}
\eta_1 \\
\eta_2
\end{pmatrix}
=
\begin{pmatrix}
\eta_1 + \eta_2 \\
\eta_2
\end{pmatrix}
$$
* If it's possible to have effects only apply to a subset of indices (?) then perhaps it could work

Warning!

\begin{enumerate}
    \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA}
    \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM
\end{enumerate}

@@

# Multiple link functions

* The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions
* For LGMs there is only one $g$, whereas ELGMs allow $g_i$
* Conclusion: \textcolor{hilit}{does} make Naomi an ELGM
* In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)`

# Statistical software

|           | Interface `y ~ 1 + x`  | General |
|-----------|------------|---------|
| Example   | `brms`     | Stan    |
| Users     | Scientists | Statisticians |
| Benefits  | Ease of use, sensible defaults, trust | Flexibility, development |
| Drawbacks | Being fenced in | Barrier to entry, failing silently |

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
\texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA}
\end{tcolorbox}
\end{center}

@@

# Moments

* Are the means and standard deviations the same? More generally the $t$th moment can be estimated by
$$
\hat{\mathbb{E}}(\theta^t) = \frac{1}{n} \sum_{i = 1}^n \theta_i^t
$$
* Not great if you're only looking at the first few moments $t = 1, 2$

# Kolmogorov-Smirov test

* Compare $F_\bullet(\vartheta)$ to $F_{\texttt{NUTS}}(\vartheta)$
$$
D_\bullet = \sup_\vartheta | F_{\texttt{NUTS}}(\vartheta) - F_\bullet(\vartheta)|.
$$
* A value $D_\bullet$ means there is at most a $(100 \cdot D_\bullet)$% difference between posterior densities

```{r, echo=FALSE, message=FALSE, fig.cap="KS test example."}
plot(1:10)
```

@@

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work,     * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Let me know if you'd be up for being an early reader!
* Code for this project is at `athowes.github.io/elgm-inf`
